---
title: "hfd_project"
author: "Isaac Hocker"
date: "2025-11-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
library(tidyverse)
library(lubridate)
# Load the data set
hfd <- read.csv("new_hfd_data.csv")
# View structure of the data set
glimpse(hfd)
```

## Introduction
This report analyzes real incident data from Houston Fire Department.
We explore trends over time, types of incidents, and identify patterns that may help with resource planning and risk prevention.

### Emt events
```{r}
ggplot(hfd, aes(x = Station, y = EMS.Events)) +
 geom_col(fill = "steelblue") +
  geom_text(aes(label = Station),
            vjust = -0.3,
            size = 3,
            color = "black") +
  labs(
    title = "EMS Events by Station",
    x = "Station Number",
    y = "EMS Events"
  ) +
  theme_minimal()
  
# This plot shows how many EMS events happen at each station
# Notes:
# we can see that there a good amount of stations that are busy with emt calls and not 
# This graph shows events 
```

#### Fire events by station
```{r}
ggplot(hfd, aes(x = Station, y = Fire.Events)) +
  geom_col(fill = "firebrick") +
  geom_text(
    aes(label = Station),
    vjust = -0.3,
    size = 3,
    angle = 90,
    check_overlap = TRUE # helps with spacing
  ) +
  labs(
    title = "Fire Events by Station",
    x = "Station Number",
    y = "Number of Fire Events"
  ) +
  theme_minimal()
# Notes:
# This graph shows how many fire related incidents each station handled.
# Higher bars indicate stations with heavier fire response activity.
# This helps identify which areas of Houston experience more frequent fire incidents.
```

##### EMS Events vs Fire Events
```{r}
ggplot(hfd, aes(x = EMS.Events, y = Fire.Events, label = Station)) +
  geom_point(color = "darkorange", size = 4) +
  geom_text(vjust = -0.5, size = 4, color = "black") +
  scale_x_continuous(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.05, 0.05)) +
  labs(
    title = "EMS Events vs Fire Events by Station",
  ) +
  theme_minimal(base_size = 14)
# Each point represents one fire station
# This plot shows whether stations with more EMS calls also tend to have more fire calls
# If the points form an upward trend it means stations with high EMS demand often also experience higher fire activity 
# also went back to scale the graph to zoom in on the station numbers
```
###### EMS Responses vs Fire Responses
```{r}
ggplot(hfd, aes(x = EMS.Responses, y = Fire.Responses, label = Station)) +
  geom_point(color = "purple", size = 3) +
  geom_text(vjust = -0.5, size = 3) +
  labs(
    title = "EMS Responses vs Fire Responses",
    x = "EMS Responses",
    y = "Fire Responses"
  ) +
  theme_minimal()
# Notes:
# For the ems vs fire responses we can see that there are way more ems incidents, but also we can see that there are certain stations with high fire response
# we also see that there are stations that get more ems responses than fire.
# If a point is going upward it shows that station has more fire responses than some of the others that go more towards the right of the graph.
```

###### Key Findings

- EMS and Fire activity are not evenly distributed across stations. Some stations handle much higher call volumes. 
- Stations with higher EMS events often also have higher fire events, suggesting certain areas have generally higher emergency demand.
- The relationship between EMS responses and fire responses indicates that resource needs may be similar across certain high demand stations.

####### Limitations
- The data set reflects only time period and may not represent long term trends.
- The data is aggregated by station and event type, so it does not include detail call level information.

######## Tools used
- Packages I used: tidyverse for data manipulation and plotting, and lubricate for working with dates
- Created bar charts to compare Ems and fire events by station
- created scatter plots to explore relationships between EMS and fire
######### What i learned
- This data gives an insight on how busy or how slow some stations are
_ Which stations also show which area or district there are located where it gives us hints that the area they are in is safer, or dangerous.
